---
title: "Activities - Scatterplot, Correlation, Crime"
author: "Wells"
date: "2025-10-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Activity #1:

Examine this New York Times scatterplot from 2013:
https://archive.nytimes.com/www.nytimes.com/interactive/2013/05/07/education/college-admissions-gap.html?_r=0

The graphic displays four different scatterplots showing relationships between student demographics and college acceptance rates. For each of the four scatterplots:

a) Describe the relationship (or lack thereof) between the two variables.

The two variables are graduation rate and percentage of freshman who are minorities at a certain college in 2011-2012. This graph shows that a lot of minority students attend colleges with low graduation rates. 

b) Note the specific words and phrases the graphic uses to characterize each relationship.

Write 1-2 sentences summarizing each scatterplot.

The first one shows how there are few black students at elite colleges. Out of the "elite" institution, UMD has a high number of black students compared to other schools like Notre Dame, Northwestern and Duke. 

The second scatterplot has a lot of clustered points to the left side. This means there are more minority students(in this graph black students) at colleges that have lower graduation rates.

The third one shows how a lot of minorities (in this case hispanics) tend to mainly go to community colleges or two year colleges. These points tend to all cluster on the left side of the graph. 

The last one shows how minorities (asians in this example) tend to make up a big proportion of the freshman class at schools that have other 80% graduation rate. For example, there tends to be a lot of asian students at "elite" schools like UCLA, Brown and NYU.

Name each person in your group who worked on this entry.

**Write Answer #1:**

-------------------------------------------------------------------------------

Activity #2: Correlation Demonstration

In the following exercise, we will look at data about 16 different consultants, their aptitude scores and their performance scores.

Our question: what is the relationship between a consultant's aptitude and their performance? 

How are they correlated?

Below we calculate a Pearson's correlation coefficient and then create a regression line.

```{r}
library(tidyverse)
```

Load data for aptitude and performance for 16 different consultants
```{r}
aptitude <- c(45, 81, 65, 87, 68, 91, 77, 61, 55, 66, 82, 93, 76, 83, 61, 74)
performance <- c(56, 74, 56, 81, 75, 84, 68, 52, 57, 82, 73, 90, 67, 79, 70, 66)
consultant <- c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16)
#this builds our table
consultants <- data.frame(consultant, aptitude, performance)
consultants
```


### Task 1: Use `summarize()` and the `cor()` function to calculate the correlation between aptitude and performance, using the pearson method.

```{r}
consultants |> summarize(correlation = cor(aptitude, performance, method="pearson"))
```

We see a Pearson's correlation coefficient of 0.783.
What does that mean?

The Pearson correlation scale is from -1 to 1.
-1 is a perfect negative correlation
1 is a perfect positive correlation

How do we interpret the values in between?

-- High Degree: Values between ±0.50 and ±1 suggest a strong correlation.
-- Moderate Degree: Values between ±0.30 and ±0.49 indicate a moderate correlation.
-- Low Degree: Values below +0.29 are considered a weak correlation.
-- No Correlation: A value of zero implies no relationship.

**Question: How would you describe the 0.783 correlation between aptitute and performance?**

**A: There is a strong correlation between aptitute and performance. Since it is close to 1, there is a strong correlation. **


### Task 2: Create a scatterplot of aptitude and performance, using `geom_point()`

```{r}

consultants |> 
  ggplot() +
  geom_point(aes(x=aptitude, y=performance)) +
  geom_smooth(aes(x=aptitude, y=performance), method="lm") +
  expand_limits(x = 0, y = 0)

```


The "best fit" line is known as a regression line that helps us understand the relationship between two variables. This predicts the dependent variable based on the independent variable. 

A regression line is a straight line showing a best-fit between independent and dependent variables.

In the example above, aptitude would be the independent variable, or the predictor. That's because we would expect aptitude to predict performance, which is the dependent variable, or the outcome variable.

Independent = predictor
Dependent = outcome

Regression is one step ahead of correlation. It helps predict the values of dependent variables or outcomes with respect to changes in independent or explanatory variables. 

Regression lines help predict outcomes.


-------------------------------------------------------------------------------

Activity #3: 

# PG Murders and Heat

Let's explore the relationship between crime and hot weather using homicide and temperature data from Prince George's County. 

You'll load the data, perform some basic statistical analysis, and create a scatterplot to explore the relationship. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

**write code below**
```{r}
library(tidyverse)
```


### Task 2: Load the dataset from [https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_murder_temps.csv](https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_murder_temps.csv)

**write code below**
```{r}
murder_temps <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_murder_temps.csv")
```



### Task 3: Use `summarize()` and the `cor()` function to calculate the correlation between murders and temperature, using the pearson method.

**write code below**
```{r}
murder_temps |> summarize(correlation = cor(murders, temperature, method="pearson"))
```



### Task 4: Create a scatterplot of murders and temperature, using `geom_point()`. Your aes() will have temperatures on the x axis, murders on the y axis and add a label for the month (so label=month inside the aes()).

**write code below**
```{r}
murder_temps |> 
  ggplot() +
  geom_point(aes(x=temperature, y=murders)) +
  geom_smooth(aes(x=temperature, y=murders), method="lm") +
  labs(title = "Relationship between the temperature and murders in PG county", x = "temperature", y ="murders")
```


### Task 5: Add labels to the scatterplot using geom_text(). Inside geom_text() put the same aes() as geom_point in Task 4, then after the aes() add size = 5.

**write code below**
```{r}
murder_temps |> 
  ggplot() +
  geom_point(aes(x=temperature, y=murders)) +
  geom_smooth(aes(x=temperature, y=murders), method="lm") +
  labs(title = "Relationship between the temperature and murders in PG county", x = "temperature", y ="murders") +
  geom_text(aes(x=temperature, y= murders, label = month))
```


### Task 6: Describe the results of both the correlation calculation and the scatterplot.

**Write Answer: There is a high correlation between the two variables because the correlation is about .6 which is considered a higher correlation(since it's between .5 and 1). Most of the months tend to fall in the close range of the line on the scatterplot which shows how the correlation is high. Although it isn't an extremely high correlation, many of the points fall close to the line and still maintain a strong correlation.


When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
