---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2024.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
library(janitor)
library(dplyr)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries<- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv") 
  
  wh_salaries |> 
  clean_names()
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

Looking at the different salaries, I can see that there is a different in most of the pays.There are some salaries that are a lot higher than others.I also noticed that there is a $0 salary for some people which makes me wonder if those people maybe get hourly paid or something. I think from a reporting standpoint, what stands out is the pay gaps and how some peoples positions get paid higher than others. Some salary values that could cause problems are that there could be missing values, outliers we don't know about or technology errors. 



### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
  wh_salaries |>
summarize(mean_salary= mean(salary),
  median_salary = median(salary))
 

```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries |> 
  summarize(mean_salary = mean(salary),
             median_salary = median(salary),
             max_salary = max (salary),
             min_salary = min (salary)
)
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

The mean shows us that the average salary is $107,036.6. The median shows us that the middle salary when you put them from lowest to highest is $110,500. Since the median is a little higher than the mean that could mean there there might be a few low salaries that are pulling the mean down. The minimum and maximum salaries tell us how there us a large gap from workers making $0-$251,258. This might mean there are unpaid interns, volunteers or just missing data since the gap is so large. The 0 is an outlier because it is so off from all the other numbers.

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

I think the best story idea would be talking about how the white house salaries range from $0 to more than a quarter of a million dollars and raising the questions of pay equity and if there are any hidden inequalities. I think investigating the outliers and interviewing those who are working for the white house right now or someone who use to. I think that could be super interesting and could be something a lot of people would want to read.
