---
title: "HW6_Spotify_Songs"
---
name : Sara Leber
In this assignment, you will be working with a dataset of songs played on Spotify in 2020, and you'll be writing code to explore the range of danceability scores in the data, using standard deviation and histograms. You'll also write up your findings.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the Spotify Songs dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv

```{r}
spotify_2020<- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv") 
```

### Task 3: Calculate the mean and standard deviation of the danceability scores

Write code to calculate the mean and standard deviation of the danceability scores in the Spotify dataset. Display those.

```{r}
spotify_2020 |>
  mutate(mean_spotify_2020 = mean(danceability),
         deviation_spotify_2020 = sd(danceability))
         
```

### Task 4: Create a histogram of the danceability scores

Write code to create a histogram of the danceability scores in the Spotify dataset, based on the example we did in class.

```{r}
spotify_2020 |>
  ggplot () +
  geom_histogram(aes(x = danceability), binwidth = 1)
```

### Task 5: Add a line to the histogram for the mean danceability score

Take the same code and add a vertical line to the histogram for the mean danceability score, based on the example in class.

```{r}
spotify_2020 |>
  ggplot () +
  geom_histogram(aes(x=danceability), binwidth = 1) +
  geom_vline(aes(xintercept = mean (danceability)), color = "red", linetype = "dashed", size = 1)
```


### Task 6: Add two lines to the histogram to show one standard deviation

Take the previous code and add two lines to the histogram for the 1st standard deviation below and above the mean score.

```{r}
spotify_2020 |>
ggplot () +
geom_histogram(aes (x = danceability), binwidth = 1) +
geom_vline (aes (xintercept = mean (danceability)), color = "red", linetype = "dashed", size = 1) +
geom_vline(aes (xintercept = mean (danceability) - sd (danceability)), color = "blue", linetype = "dashed", size = 1) +
geom_vline(aes (xintercept = mean (danceability) + sd(danceability)), color = "blue", Linetype = "dashed", size = 1)
```

### Task 7: Describe our results and pitch a story

Write a few sentences about the distribution of danceability scores in the Spotify dataset. Is it a normal distribution, or is it skewed in one direction? What does the mean and one standard distribution tell us about the songs in the dataset? Draft a short story pitch from this data.

The danceability scores in the Spotify's dataset is skewed towards the right, meaning most songs on Spotify are clustered toward the higher values. The distriution of danceability shows that Spotify songs tend to be high danceable and there tends to not be many songs that aren't danceable.  The mean is slightly below the median which mean there is a fewer but more spread out values on the lower side. One standard deviation above and below the mean shows where 68%(or around two out of three) of all songs fall if the distruction is normal and the mean is below the big peak of songs which shows that some of the low danceability songs are pulling it towards the left. I think a good story about this topic would be about how Spotify is made for the dance floor and has high rates of danceability. I think it could be interesting if you compared the numbers of danceability to other platforms like Apple Music or Sound Cloud and then interview people on which one they would think would be the highest and why. For the story the writer could also dive deeper in the songs on Spotify and why these songs seem to mostly be songs you can easily dance too.

### Task 8: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
