---
title: "Data Distribution, Volleyball Percentiles, Car Thefts, "
author: "Derek Willis /Rob Wells"
date: 09-23-2025
---

This workbook contains three exercises for our class. Complete them and hand it in at the end of class via a GitHub link.

Your name:
Sara Leber


#------------------------------------------------------------------------------#


# Part 1: When Data Isn't Normal

Working in pairs, consider the following question: "Why do many statistical methods assume a normal distribution, and what challenges arise when data does not meet this assumption?"

First, come up with two examples of data that might not be normally distributed - in other words, that could be skewed or unequal in some way - and explain why they wouldn't be normally distributed. Feel free to draw what you think the distribution would look like.

Then, how should a reporter working on a story about data that isn't normally distributed describe that data? Which statistical measures that we've covered in class should they use or avoid when doing so? Use the examples you came up with, and you should discuss your answers together, but each student submits separately.

**Your answer**: An example of data that might not be normally distributed is test scores because there tends to be all different scores on a test and wide range of test scores.It can between a 05 and 100%. Another example is the income in a household because there usually is a wide variety in 
incomes from jobs especially in the world right now.

If a reporter is working on a story that the data might not be normally distributed, the journalist should be clear that the data is skewed and try to find out the change in the data rates and adjust to the distorted data.Find the outliers and try to find the median not the mean.




#------------------------------------------------------------------------------#


# Part 2: Volleyball Percentiles

We'll be calculating percentiles to see how good Maryland was in service aces in the 2024 season. We'll use `mutate()` and a new function `ntile()` to calculate percentiles. Get started by loading the tidyverse in a code block, then load the CSV file here: https://raw.githubusercontent.com/dwillis/NCAAWomensVolleyballData/refs/heads/main/data/ncaa_womens_volleyball_matchstats_2024.csv" and save it to a dataframe called `matches`.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the data

```{r}
matches <- read_csv("https://raw.githubusercontent.com/dwillis/NCAAWomensVolleyballData/refs/heads/main/data/ncaa_womens_volleyball_matchstats_2024.csv")
```

### Task 3: Create totals of aces for each team using `group_by()` and `summarize()`

```{r}
team_aces <- matches |> 
  group_by(team) |> 
  summarize(total_aces = sum(aces)) |> 
  arrange(desc(total_aces))
```


### Task 4: Calculate percentiles using `mutate()` and `ntile()`

```{r}
team_aces |> mutate(percentile = ntile(total_aces, 100))
```


### Task 5: Describe the percentiles

Write a couple of sentences that puts Maryland into context with the other teams in the dataset. You can do some Internet research to help with this.

**Answer**:
Compared to other teams amount of aces, Maryland is right in there with big, well-known teams and is in the 95th percentile. The percentile that Maryland is in the top 20th in ranking for aces shows they are a strong serving team compared to other teams.  Other teams, have scored only 5% more than Maryland.


#------------------------------------------------------------------------------#

# Part 3: Car Thefts

Follow along with the demo to learn how to calculate standard deviation and how to create a histogram of the data using R and the Tidyverse. Get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```


### Task 2: Load the car thefts dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_jan24.csv


```{r}
car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_jan24.csv")
```

### Task 3: Calculate the standard deviation of the number of car thefts, along with the mean

```{r}
car_thefts |> summarize(mean = mean(count), sd = sd(count))
```


### Task 4: Create a histogram of the number of car thefts


```{r}
car_thefts |> 
  ggplot() + 
  geom_histogram(aes(x = count), binwidth = 2)
```




### Task 5: Add a line to the histogram that shows the mean of the number of car thefts, and two other lines to show one standard deviation in both directions

```{r}
car_thefts |> 
  ggplot() + 
  geom_histogram(aes(x = count), binwidth = 2) +
  geom_vline(aes(xintercept = mean(count)), color = "red", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(count) - sd(count)), color = "blue", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(count) + sd(count)), color = "blue", linetype = "dashed", size = 1)
```

### Task 6: Describe your results

Is the distribution normal or skewed in one direction? Write a sentence in which you describe the range of the data covered by one standard deviation from the mean on both sides - how much of the data falls in that range?


**Answer**:
The data is skewed to the right. The standard deviation of the mean is 4.3  while the mean is around 12.7. Although the mean is around 12.7, there is some variation seen and the data could fall between 4.3 below or above 12.7.



When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.

Congrats!


